Finding New Pulsars Using Machine Learning

### 2020-05-25 Draft.tex
methods

% Coding packages
\usepackage{listings}
\usepackage{lmodern}
\usepackage{xcolor}
\lstset{
  basicstyle=\ttfamily,
  literate={~} {\raise.17ex\hbox{$\scriptstyle\mathtt{\sim}$}}{1},
  columns=fullflexible,
  frame=single,
  breaklines=true,
  numbers=left,
  stepnumber=1,
  xleftmargin=2em,
  framexleftmargin=1.5em
}

@@

% Create the Title 
\title{Finding New Pulsars Using Machine Learning}
\author{Jacob Ian Matthews \\ \\ Draft 1}

@@

The Machine Learning Training Dataset has now been created using Murchison Widefield Array data.
\\

@@

@@

The machine learning training dataset created with candidates detected by the Murchison Widefield Array is as follows:

\begin{lstlisting}[title=trainingSet.arff]

\end{lstlisting}

@@

\pagebreak
% Results Discussion Section
\section{Discussion}
\subsection{How effective are the training datasets?}
\subsection{Why didn't the LOFAR Machine Learning Algorithm work with the MWA?}
\subsection{What changes to the algorithm were necessary for it to successfully classify Pulsars from the MWA?}

@@

>>>>

% Coding packages
\usepackage{listings}
\usepackage{lmodern}
\usepackage{xcolor}
\usepackage{algorithmicx}
\lstset{
  basicstyle=\ttfamily,
  literate={~} {\raise.17ex\hbox{$\scriptstyle\mathtt{\sim}$}}{1},
  columns=fullflexible,
  frame=single,
  breaklines=true,
  numbers=left,
  stepnumber=1,
  xleftmargin=2em,
  framexleftmargin=1.5em
}

@@

% Create the Title 
\title{Finding New Pulsars Using Machine Learning}
\author{Jacob Ian Matthews \\ \\ Draft}

@@

The Machine Learning Training Dataset has now been created using Murchison Widefield Array data.

@@

To address the Candidate Selection Problem discussed in \emph{Section 1.3} of this report, a Java Machine Learning pulsar classification tool named \verb|LOTAASClassifier| was created to classify the pulsar candidates produced by the LOFAR Tied-Array All-sky Survey (LOTAAS) \autocite{lyon}. Due to the similarities between the LOFAR radio telescope and the Murchison Widefield Array (MWA) radio telescope, it is logical to attempt to apply the \verb|LOTAASClassifier| tool to candidates produced by the MWA.

The \verb|LOTAASClassifier| tool contains four machine learning algorithms with which the user can choose to make pulsar classifications. They are: the J48 (C4.5 Decision Tree \autocite{quinlan}), the Multi-Layer Perceptron, the Naive Bayes, and the Support Vector Machines (SVM) algorithms \autocite{lyon}.

To begin evaluating the software, \verb|LOTAASClassifier v1.0| can be downloaded from its GitHub repository \\(\verb|https://github.com/scienceguyrob/LOTAASClassifier|) with the following UNIX terminal commands:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars
$ git clone https://github.com/scienceguyrob/LOTAASClassifier.git
\end{lstlisting}

The tool's executable program can then be found by navigating to the following directory in the UNIX terminal:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars/LOTAASClassifier/dist
\end{lstlisting}

\subsubsection{Creating a Classification Model}

In order to use \verb|LOTAASClassifier| to classify Murchison Widefield Array candidates, we must use the Machine Learning Training Dataset created in \emph{Section 2.1} to create a classification model. This can be completed by running the following commands:

\begin{lstlisting}[numbers=none]
$ java -jar LOTAASClassifier.jar -t ~/pulsars/data/trainingSet.arff -m ~/pulsars/data/model.m -a 1
\end{lstlisting}

This instructs \verb|LOTAASClassifier| to create a new classification model for the J48 machine learning algorithm with the previously constructed training dataset.

To ensure that the classification model was created successfully, we can test the model against the same dataset provided by N. Swainston to create the training dataset. Using \verb|PulsarFeatureLab| we can produce a new \verb|output.arff| file with the candidates' class unedited.

It is now possible to test the classification model with the candidate feature sets with the following commands:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars/LOTAASClassifier/dist
$ java -jar LOTAASClassifier.jar -p ~/pulsars/data/pfd/output.arff -m ~/pulsars/data/model.m -a 1
\end{lstlisting}

The \verb|LOTAASClassifier| tool will output two files: `\verb|output.positive|' and `\verb|output.negative|', in the same location as the input dataset. The J48 classification model was constructed successfully if the candidates inside the `\verb|positive|' file are the known pulsars.

\subsubsection{Evaluating LOTAASClassifier}

With a successfully created machine learning classification model, it is now possible to test \verb|LOTAASClassifier| against a previously unseen (by the classifier) dataset of Murchison Widefield Array candidates whose class is also known. We can first remove the existing candidates from inside the \verb|PFD| files directory and delete the existing \verb|output| files.

The candidates directory can then be populated by new \verb|PFD| files and we can use the `\verb|docker run|' command from \emph{Section 2.2.1} to extract the features from the new candidates.

The \verb|LOTAASClassifier| tool can now make classification predictions on the new candidates by running the UNIX terminal command:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars/LOTAASClassifier/dist
$ java -jar LOTAASClassifier.jar -p ~/pulsars/data/pfd/output.arff -m ~/pulsars/data/model.m -a 1
\end{lstlisting}

By inspecting the created `\verb|output.positive|' and `\verb|output.negative|' files and comparing the results to the known pulsars, we can evaluate the performance of the \verb|LOTAASClassifier| tool with Murchison Widefield Array data and make a judgement about further use of the classifier.

\subsubsection{Creating the PulsarClassifier Ensemble Classification Tool}

According to \cite{tan}, using Machine Learning algorithms in ensemble to make pulsar classifications increases the accuracy of classifications, classifying pulsars that were often misclassified such as wide-pulse pulsars.

To build the ensemble classification feature into the existing \verb|LOTAASClassifier| tool, we can first begin by creating a new Java project named \verb|PulsarClassifier| using the free software, Maven (\verb|https://maven.apache.org|).

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars
$ mkdir PulsarClassifier && cd PulsarClassifier
$ mvn archetype:generate -DarchetypeGroupId=org.apache.maven.archetypes -DarchetypeArtifactId=maven-archetype-quickstart -DarchetypeVersion=1.4
\end{lstlisting}

We can now copy the source code from \verb|LOTAASClassifier| to be included in the \verb|PulsarClassifier| software.

\begin{lstlisting}[numbers=none]
$ cp -R ~/pulsars/LOTAASClassifier/src/ ~/pulsars/PulsarClassifier/src/main/java
\end{lstlisting}

To use the WEKA suite of Machine Learning tools, we must then edit the \verb|pom.xml| file inside \verb|PulsarClassifier| to include it as a dependency, and resynchronise the project:\\

\begin{lstlisting}[numbers=none, title=pom.xml, language=xml]
...
<dependencies>
...
    <dependency>
        <groupId>nz.ac.waikato.cms.weka</groupId>
        <artifactId>weka-stable</artifactId>
        <version>3.8.0</version>
    </dependency>
...
</dependencies>
...
\end{lstlisting}

We now have the following basic project directory structure:

\begin{lstlisting}[numbers=none]
PulsarClassifier/
    src/
        main/
            java/
                com/jacobianmatthews/pulsarclassifier/
                com/scienceguyrob/lotaasclassifier/
        test/
            java/
                com/jacobianmatthews/pulsarclassifier
    target/
            ...
    pom.xml
\end{lstlisting}

Where the new source code will be located under \\\verb|/src/main/java/com/jacobianmatthews/pulsarclassifier|. To introduce the ensemble classification feature, we must write four main Java classes: \\\verb|PulsarClassifier.java|,  \verb|ClassifierBuilder.java|,  \verb|ClassifierValidator.java|,  and \verb|ClassPredictor.java|.

The \verb|LOTAASClassifier| tool accepts a command-line argument \verb|-a| which accepts an integer that denotes the machine learning algorithm to use in building a classification model and making predictions \autocite{lyon}. Therefore, we will add an algorithm into the above listed Java classes that will accept an integer value of \verb|-1| that will activate the ensemble classifier.

To add ensemble classification to the class \verb|ClassifierBuilder.java|, we will use the following algorithm:

\begin{algorithmic}[1]

\end{algorithmic}

\subsubsection{Evaluating PulsarClassifier}

@@

The machine learning training dataset created with candidates detected by the Murchison Widefield Array can be found under \emph{Appendix 8.3}.

@@

\pagebreak
% Results Discussion Section
\section{Discussion}
\subsection{How effective are the training datasets?}
\subsection{Evaluating Curtin Institute of Radio Astronomy's Pulsar Classification Pipeline}
\subsection{Why did/didn't the LOFAR Machine Learning Algorithm work with the MWA?}
\subsection{What changes to the algorithm were necessary for it to successfully classify Pulsars from the MWA?}

@@

\pagebreak
\subsection{Machine Learning Training Dataset}
\lstinputlisting[title=trainingData.arff, basicstyle=\footnotesize\ttfamily, breaklines=true]{code/trainingSet.arff}
